---
title: "Confidence Envelope"
author: "Doug Nychka"
date: "`r Sys.Date()`"
output: pdf_document
---

## Some setup 
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
setwd(
  "~/Dropbox/Home/Teaching/FDA/theCourse/Modules/Mod03LeastSquaresSmoothing"
  )
suppressMessages(library(fda))
suppressMessages( library(fields))
```

Loading in a handy R function to create the natural soline basis 
```{r}
source("../R/naturalSplineFit.R")
source("../R/naturalSplineBasis.R")
```

## Synthetic data example

```{r}
# synthetic dataset
set.seed(123)
N <- 150
s <- runif(N)
s <- sort(s)
true <-  9 * s * (1 - s) ^ 3
# rnorm generates standard normals 
# multiply by .1 to get sigma of .1
errors<- .1 * rnorm(N)
y <-  true + errors
plot( s,y)
lines( s, true, col="magenta", lwd=2)
title("True curve with data")
```

\newpage

# Fitting with natural spline and 7 knots 

Natural splines are as many as knots due to the
  BC at either end ( second deriv. == 0).
  
This should be a easy application of the **lm** function. However, 
I lost an hour figuring out how to get lm to use a basis matrix and 
then predict!
The   ```.```   in the code below means use all the variables in 
the data frame.  Both **Phi** and **PhiGrid** will have the same columns names because they are converted to data frames the same way. 
  
```{r}

sKnots <- seq(0, 1, length.out = 7)

# make sure you understand how this function 
# works

Phi <-  naturalSplineBasis(s, sKnots)
# omit intercept because it is already included as part of the
# basis
Phi<- data.frame( Phi)


fit <- lm(y ~ .-1, data=Phi  )
summary( fit)
```


# prediction intervals on grid 

```{r}
sGrid <- seq(0, 1, .02)
PhiGrid <- naturalSplineBasis(sGrid, sKnots)
# trick is that df gets the same variable names as those in the fit object.

df<- data.frame(PhiGrid)

out<- predict( fit, df  ,
               se.fit=TRUE )
 
 SE<- out$se.fit
 fHat<- out$fit

P<- length( sKnots)
GCV <- (1 / N) * sum(fit$residuals ^ 2) / (1 - P / N) ^ 2
```

# bonferroni Z

Use ```lower.tail=FALSE``` to find the upper tail probability
this just makes this easier to read. Compare **zB** below  to 1.96

```{r}
  zB<- qnorm( .025/length( sGrid), lower.tail = FALSE)
  zB
```

  
\newpage

# Envelope plot 

Here is a way to create a nice  envelope plot. It uses the basic graphics function **polygon**.

```{r}
plot(s, y, pch=16, cex=.5, type="n")
envelopePlot(sGrid,fHat - zB*SE,sGrid, fHat + zB*SE )
points(s, y ,pch=16, cex=.5)
lines(s, true, lwd = 2, col = "black")
# fHat 
lines(sGrid, fHat, col = "red", lty = 1)
```
